---
title: "Covid-19 Project"
author: "Siavash Keivani"
date: "2023-03-01"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

### COVID-19 Data Analysis

In this analysis we will take a look at USA national covid-19 cases and deaths and 
compare them to trends in California. We will also create 2 models to asses the 
impact of cases per thousand on deaths per thousand, and to predict USA deaths 
per million in 2024 using the open source prophet library.

These data sets are provided by Johns Hopkins on Github. This data is gathered
from different sources which are listed by Johns Hopkins on their github links 
below.

This data set includes location information for countries, and the number of
covid cases and deaths. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
options(width=120)
library("tidyverse")
library("lubridate")
library("dplyr")
library("ggplot2")
library("scales")
library("readr")
library("prophet")
library("sf")
library("ggspatial")
library("plotly")
library("maps")
library("usmap")
```


##  Libraries Used


library("tidyverse")
library("lubridate")
library("dplyr")
library("ggplot2")
library("scales")
library("readr")
library("prophet")
library("sf")
library("ggspatial")
library("plotly")
library("maps")
library("usmap")


## Read in Data

```{r}
urlfile1 <- 'https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv'
urlfile2 <- 'https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv'
urlfile3 <- 'https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv'
urlfile4 <- 'https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv'
```

```{r}
us.cases <- read_csv(url(urlfile1))
global.cases <- read_csv(url(urlfile2))
us.deaths <- read_csv(url(urlfile3))
global.deaths <- read_csv(url(urlfile4))
```
## Tidy Data

Tidy global.cases

```{r}
# Pivot global cases and name columns
global.cases <- global.cases %>%
  pivot_longer(cols = -c('Province/State','Country/Region','Lat','Long'),
    names_to = "date",
    values_to = "cases")
```

```{r}
#remove lat long columns
global.cases<-select(global.cases,-c(Lat,Long))
```

Tidy global.deaths

```{r}
# pivot and give column names
global.deaths <- global.deaths %>%
  pivot_longer(cols = -c('Province/State','Country/Region','Lat','Long'),
    names_to = "date",
    values_to = "deaths")
```

```{r}
# remove lat and long columns
global.deaths<-select(global.deaths,-c(Lat,Long))
```

```{r}
# Merge global cases and deaths
global <- global.cases %>%
  full_join(global.deaths) %>%
  rename(Country_Region = 'Country/Region', Province_State = 'Province/State') %>%
  mutate(date = myd(date))
```
```{r}
summary(global)
```
```{r}
# only cases above 0
global<-global %>% filter(cases > 0)
```

## Wrangling, cleaning, and merging us.deaths and us.cases

```{r}
us.cases <- us.cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date))%>%
  select(-c(Lat,Long_))
```

```{r}
us.deaths <- us.deaths %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date))%>%
  select(-c(Lat,Long_))
```

```{r}
us <- us.cases %>%
  full_join(us.deaths)
```
## Data wrangling for global data sets

```{r}
global <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ",",
        na.rm = TRUE,
        remove = FALSE)
```

```{r}
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
```

```{r}
uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2 ))
```

```{r}
global <- global %>%
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS))%>%
  select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)
```


```{r}
us.state <- us %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths*1000000 / Population) %>%
  select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()
```
## Creating a Map of Deaths per US State

```{r}
us.state.md <- us.state %>%
  select(Province_State,deaths,cases,Population) %>%
  rename(state = "Province_State") %>%
  group_by(state) %>%
  summarize(deaths = max(deaths),cases =  max(cases), Population = max(Population)) %>%
  mutate(deaths.pc=deaths/Population, cases.pc = cases/Population )
 # rename(state = "Province_State")
```

```{r}
plot_usmap(data = us.state.md, values = "deaths", regions = "states") +
  
  labs(title = "US States",
       subtitle = "Map of Covid Deaths per US State ") + 
  theme(panel.background = element_rect(color = "black", fill = "lightblue")) +
  scale_fill_continuous(
    low = "yellow", high = "red", name = "deaths", label = scales::comma) + 
  theme(legend.position = "right")
```



## Map of Cases per US State


```{r}
plot_usmap(data = us.state.md, values = "cases", regions = "states") +
  
  labs(title = "US States",
       subtitle = "Map of Covid Cases per US State ") + 
  theme(panel.background = element_rect(color = "black", fill = "lightblue")) +
  scale_fill_continuous(
    low = "white", high = "blue", name = "cases", label = scales::comma) + 
  theme(legend.position = "right")
```

The maps above show California, Texas and Florida as being states with large
amounts of cases and deaths. However a more accurate analysis might be to look
at cases and deaths per capita. Lets take a look below.

```{r}
plot_usmap(data = us.state.md, values = "cases.pc", regions = "states") +
  
  labs(title = "US States",
       subtitle = "Map of Covid Cases per Capita by US State ") + 
  theme(panel.background = element_rect(color = "black", fill = "lightblue")) +
  scale_fill_continuous(
    low = "white", high = "blue", name = "cases per capita", label = scales::comma) + 
  theme(legend.position = "right")
```

```{r}
plot_usmap(data = us.state.md, values = "deaths.pc", regions = "states") +
  
  labs(title = "US States",
       subtitle = "Map of Covid Deaths per Capita by  US State ") + 
  theme(panel.background = element_rect(color = "black", fill = "lightblue")) +
  scale_fill_continuous(
    low = "yellow", high = "red", name = "deaths per capita", label = scales::comma) + 
  theme(legend.position = "right")
```

Cases and deaths per capita show a different story when it comes to CA, TX, and
FL. California and Texas seem to now be in the  middle of the pack in contrast 
to before when the maps showed them in the lead with cases and deaths.


```{r}
us.totals <- us.state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths*1000000 / Population) %>%
  select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()
```

```{r fig.width= 15, fig.height= 7}
us.totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color="cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10()+
  theme(legend.position = "bottom", axis.text.x = element_text(angle=90)) +
  labs(title = "COVID19 in US", y = NULL)
  
```

```{r fig.width= 15, fig.height= 7}
state <- "California"
us.state %>%
  filter(Province_State == state) %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color="cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10()+
  theme(legend.position = "bottom", axis.text.x = element_text(angle=90)) +
  labs(title = "COVID19 in California", y = NULL)
  
```
We see from the above visualizations that covid cases and deaths follow similar
trends when comparing California to the USA as a whole. We see an exponential 
rise in both cases and deaths. then in 2022-2023 we see that cases and deaths 
plateu. Early on deaths were closer to cases but as vaccines and new treatments
rolled out it seems that the gap between deaths and cases has increased with many
more cases than deaths.


## Taking a Look at New Cases and New Deaths

```{r}
# creating new cases and deaths columns
us.state <- us.state %>%
  mutate(new_cases = cases - lag(cases), new_deaths = deaths - lag(deaths))
us.totals<- us.totals %>%
  mutate(new_cases = cases - lag(cases), new_deaths = deaths - lag(deaths))
```

## Visualizing New Cases and New Deaths

```{r fig.width= 10, fig.height= 7}
state <- "California"
us.state %>%
  filter(Province_State == state) %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color="new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10()+
  theme(legend.position = "bottom", axis.text.x = element_text(angle=90)) +
  labs(title = "New COVID19 Cases & Deaths in California", y = NULL) +
  geom_smooth(aes(x=date,y=new_deaths),method = loess) +
  geom_smooth(aes(x=date,y=new_cases), method = loess)
  
  
```
```{r fig.width= 10, fig.height= 7}
state <- "Texas"
us.state %>%
  filter(Province_State == state) %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color="new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10()+
  theme(legend.position = "bottom", axis.text.x = element_text(angle=90)) +
  labs(title = "New COVID19 Cases & Deaths in Texas", y = NULL) +
  geom_smooth(aes(x=date,y=new_deaths), method = loess) +
  geom_smooth(aes(x=date,y=new_cases), methos = loess)
  
```


```{r fig.width= 10, fig.height= 7}
us.totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color="new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10()+
  theme(legend.position = "bottom", axis.text.x = element_text(angle=90)) +
  labs(title = "New COVID19 Cases & Deaths in US", y = NULL) +
  geom_smooth(aes(x=date,y=new_deaths), method = loess) +
  geom_smooth(aes(x=date,y=new_cases), methos = loess)
  
```

Data in 2022-2023 is showing a slight decline in new cases and new deaths. We have to
think about why this might be the case. It is possible that this is actually 
the case however data reporting during  this  time might be a lot weaker.

During the height of the pandemic people were more likely to report their covid 
case. Now in a world of vaccines and boosters it is quite possible that people
do not report when they test positive especially with the large amount of at 
home test available to the public.


```{r}
us.state.totals <- us.state %>%
  group_by(Province_State)%>%
  summarize(deaths = max(deaths), cases = max(cases), population = max(Population),
            cases_per_thou = cases*1000/population, deaths_per_thou = deaths*1000/population) %>%
  filter(cases > 0, population > 0)
```

```{r fig.height= 15, fig.width=5}
us.state.totals %>%
  ggplot(aes(x = Province_State, y = cases_per_thou))+
  geom_bar(stat="identity", fill='steelblue')+
  coord_flip()+
  theme_minimal()
```




A glance at cases per thousand show most states between 200 and 350.



Taking a look at the highest and lowest 10 states for deaths per thousand

```{r}
us.state.totals %>%
  slice_min(deaths_per_thou, n =10)
```
```{r}
us.state.totals %>%
  slice_max(deaths_per_thou, n =10)
```

## Linear model of deaths per thousand as a function of cases per thousand.
```{r}
lm1 <- lm(deaths_per_thou ~ cases_per_thou, data = us.state.totals)
summary(lm1)
```
The low p value shows that cases per thousand is statistically significant when
predicting deaths per thousand in the united states.

```{r}
plot(lm1)
```

Our residual vs fitted plot indicates that the data might not be linear however
this does not mean we cant gain any insight from our model.

## Visualizing cases per thousand and deaths per thousand

```{r}
ggplot(us.state.totals, aes(x=cases_per_thou, y=deaths_per_thou)) +
  geom_point(size=1.5, shape=1)+
  geom_smooth(method='loess', formula= y~x)
```




While it may seem that the relationship is not linear we have to ask why that might be?
As stated before it is possible that as time passes, and the vaccines 
and treatments are rolled out, that people are less likely to report a positive 
test, especially given the abundance of at home test. 

```{r}
ggplot(us.totals, aes(x=cases, y=deaths)) +
  geom_point(size=1.5, shape=1) +
  geom_smooth(method='lm', formula= y~x)
```






## Creating a Model to Predict Deaths per Million in the USA

using an open source library called prophet which generates a model used to
predict future outcomes.

```{r}
# format data for use in the prophet function
us.totals.m <- us.totals %>%
  select(date, deaths_per_mill)%>%
  rename(
    ds = date,
    y = deaths_per_mill
    )
```

```{r}
library(prophet)
mod <- prophet(us.totals.m, yearly.seasonality = TRUE, daily.seasonality = TRUE)
```

```{r}
pred <- make_future_dataframe(mod,periods = 365)
fc<-predict(mod,pred)
plot(mod,fc)
```




This model seems to predict a rise in deaths per million in 2024 as compared to 
previous years. However, this is not accounting for better treatments, and more
effective vaccines / booster. 



## Bias Identification

COVID19 was a very hot political topic in the USA and as a US citizen I certainly 
had some bias regarding this topic. I had initially thought states like California,
Oregon, New York, and Washington would have the worst performance when it comes
to controlling the spread of the virus and the death counts. Looking at raw state
deaths and cases initially confirmed by bias, however a further analysis using maps
and per capita calculations showed that in fact those mentioned states were not
the worst performing. 

Another potential source of bias is in the data capturing process. With the
potential for less self reporting as time goes on, the data that is  reported might
only be of those whos cases were severe enough to be hospitalized or see a doctor.
As we know, a very large portion of folks who are vaccinated and otherwise healthy
will most likely not reach this state and therefore their case might not be captured.
It is hard to tell exactly how many cases this would be but it is likely enough to
have an effect on conclusions made based on our calculations.

## Conclusion

Our analysis showed that new cases and deaths seem to have platued across the USA.
Comparing this to California we can see a similar trend. We also were able to 
determine that while TX, CA, and FL were the highest in raw deaths and cases,
TX and CA specifically were actually lower when it came to cases and deaths per 
capita when compared to the majority of US states.

We were able to show using a linear model that cases per thousand was statistically
significant when predicting deaths per thousand. 

Our 2nd model predicted a potential increase in deaths per million in the USA in 2024
when compared to 2023. This is with the caveat that this cannot factor in the potential
for more effective treatments and vaccines/boosters.




















