---
title: "NYPD Shooting by Age and Location"
output:
  html_document: default
  pdf_document: default
date: "2023-02-08"
---
### Purpose and Problem Statement

In this analysis we will take a look at the nypd shooting data and attempt
to ascertain what age groups are more responsible for shootings and if factors
such as location descriptions, age group, and boro are a significant predictor 
of a shooting being a murder.

This information could potentially be used to impact policy designed to reduce 
shootings / murders.


Below are the list of packages required to run this file.

library("tidyverse")
library("lubridate")
library("dplyr")
library("ggplot2")
library("scales")


```{r setup}
knitr::opts_chunk$set(echo = FALSE)
options(width = 120)
library("tidyverse")
library("lubridate")
library("dplyr")
library("ggplot2")
library("scales")
```

Downloading the data.
```{r data}
url_in <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
file_names<-c("NYPD Shooting Incident Data (Historic)")
```

Reading the csv file from the above link.
```{r import data, echo=TRUE}
nypd<-read_csv(url_in[1])
```

Cleaning Data.

```{r sumamry, echo=TRUE}
#Tidy data and give summary
#Remove columns not needed
nypd<-select(nypd, -c(JURISDICTION_CODE, Lon_Lat, X_COORD_CD, Y_COORD_CD, Latitude, Longitude, BORO, PRECINCT, PERP_RACE, ))

#Format OCCUR_DATE Column
nypd$OCCUR_DATE<-as.Date(nypd$OCCUR_DATE, "%m/%d/%Y")

#Omit all NA values for locations
nypd.l<-subset(nypd, !is.na(nypd$LOCATION_DESC))
```

```{r summary, echo=TRUE}
summary(nypd.l)
```

```{r}
nypd.l <- nypd.l %>% filter(PERP_AGE_GROUP != '224' & PERP_AGE_GROUP != '1020' & PERP_AGE_GROUP != '940' )
nypd <- nypd %>% filter(PERP_AGE_GROUP != '224' & PERP_AGE_GROUP != '1020' & PERP_AGE_GROUP != '940' )
```


## Ages of Shooters and Victims

```{r}
nypd.perp.c <- nypd.l %>% count(PERP_AGE_GROUP)
colnames(nypd.perp.c) <- c('age','count.perp')
```

```{r}
nypd.vic.c<-nypd.l %>% count(VIC_AGE_GROUP)
colnames(nypd.vic.c) <- c('age','count.vic')
```

```{r}
nypd.c <-merge(nypd.perp.c,nypd.vic.c, by = 'age')
nypd.c
```
The above table shows that the majority of known perps are between the ages of 
18 and 44.

The table also shows that the majority of known victims are also between the 
ages of 18 and 44 with a larger amount in the 25-44 year old group when compared to 
perpetrators.


### Locations of Shootings

Below we analyze the locations of shootings then conduct the same analysis based
on age groups.

```{r fig.width=15, fig.height=9}
ggplot(nypd.l, aes(x=LOCATION_DESC)) + geom_bar(stat='count',fill = 'steelblue')+
  coord_flip()+
  geom_text(stat = 'count',aes(label=after_stat(count)), hjust = -.01) +
  theme_minimal() +
  ggtitle("Shootings by Location")
```

Above is a brief view into the different locations where shootings occurs. 
My goal is to take this insight and analyze the locations in relation to victim 
and perpetrator ages. Essentially, where do young people commit shootings / 
are victims and where do older people commit shootings / are victims?


# Age Based Analysis of Shooting Locations

```{r}
#taking a look at ages <18 and 18-24 to see where younger victims are impacted by crime the most
nypd.young.vic<-nypd.l
nypd.young.vic<-nypd.young.vic[nypd.young.vic$VIC_AGE_GROUP == c("<18","18-24","25-44"),]
```

```{r}
nypd.young.vic.count<- nypd.young.vic%>%
  dplyr::count(LOCATION_DESC)%>%
  dplyr::mutate(perc = n/sum(n) * 100)

```

```{r}
#taking a look at ages 45-64 and 64+ to see where  victims are impacted by crime the most
nypd.old.vic<-nypd.l
nypd.old.vic<-nypd.old.vic[nypd.old.vic$VIC_AGE_GROUP == c("45-64","65+") ,]

```

```{r fig.width=15, fig.height=9}
ggplot(nypd.young.vic.count, aes(x = LOCATION_DESC, y = n),) +
  geom_col(fill = 'steelblue') +
  geom_text(aes(x = LOCATION_DESC, y = n
                , label = paste0("","" , round(perc,2),"%")
                , hjust = -.02
                )) +
  coord_flip() +
  ggtitle("Shooting Locations Victims ages <18-44")+
  theme_minimal() 
```

Above is a visual that shows the percentage of locations of reported crimes
for a substet of the nypd dataset that includes only victims between the ages
of <18 and 44. We can see that most (67.6%) of young victims are located in
multi-dwell public housing and apartment buildings. Grocery stores and bars/
nightclubs make up 11.51% of the locations as well. Will view similar data but
for older victims below.


```{r}
# creating a frequency / percentage table from nypd.old

nypd.old.vic.count<- nypd.old.vic%>%
  dplyr::count(LOCATION_DESC)%>%
  dplyr::mutate(perc = n/sum(n) * 100)
```

```{r fig.width=15, fig.height=9}
ggplot(nypd.old.vic.count, aes(x = LOCATION_DESC, y = n),) +
  geom_col(fill = 'steelblue') +
  geom_text(aes(x = LOCATION_DESC, y = n
                , label = paste0("","" , round(perc,2),"%")
                , hjust = -.02
                )) +
  coord_flip() +
  ggtitle("Shooting Locations Victims Ages 45-65+") +
  theme_minimal()
```



At a brief glance it seems that like younger victims, older victims tend to 
have the shootings impact them in multi dwell public housing and apartment buildings.
While this is true, the percentages tell a slightly different story. In
comparison to the ages of <18-44, victims between the ages of 44-65+ show a 18.43%
decrease in shootings in multi dwell housing and apartments. 

Overall the other locations which still contain significantly less amounts of
shootings in comparison to multi dwell public housing and apartments 
seem to be more represented in the older victim group. For example
private homes, social clubs, and all the  different kinds of 
stores seem to be more represented in older victims. However Bars and nightclubs
are less represented.


## Potential for Bias from NA Values

Perpetrator data for both age groups (<18-44, and 44-65+) contain majority
na values. This could be for a multitude of reasons. Not every criminal
is caught so the missing values could represent those cases, however this is not 
conclusive. For further analysis we will be removing na values from both data 
sets but this might introduce a risk of survivor bias.


## Format & Tidy Data for a Logistic Regression Models

```{r}
# Read in CSV from link
nypd.1 <- read_csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD") 
```

```{r}

# Select appropriate columns and tidy data

nypd.m<-nypd.1
#nypd.m <-drop_na(nypd.m)
nypd.m<-nypd.m %>% select(INCIDENT_KEY, 
                   OCCUR_DATE,
                   OCCUR_TIME,
                   BORO, 
                   STATISTICAL_MURDER_FLAG,
                   PERP_AGE_GROUP,
                   PERP_SEX,
                   PERP_RACE,
                   VIC_AGE_GROUP,
                   VIC_SEX,
                   VIC_RACE,
                   LOCATION_DESC
                  )

nypd.m <- nypd.m %>% filter(PERP_AGE_GROUP != '224' & PERP_AGE_GROUP != '1020' & PERP_AGE_GROUP != '940' & PERP_AGE_GROUP != 'UNKNOWN')
nypd.m<-nypd.m %>% filter(VIC_SEX != "U")
nypd.m<-nypd.m %>% filter(PERP_SEX != "U")
nypd.m<-nypd.m %>% filter(VIC_AGE_GROUP != "UNKOWN")

nypd.m$STATISTICAL_MURDER_FLAG[nypd.m$STATISTICAL_MURDER_FLAG == "TRUE"] = 'a'
nypd.m$STATISTICAL_MURDER_FLAG[nypd.m$STATISTICAL_MURDER_FLAG == "FALSE"] = 'b'

# Format as factor

nypd.m$BORO <-as.factor(nypd.m$BORO)
nypd.m$LOCATION_DESC <-as.factor(nypd.m$LOCATION_DESC)
nypd.m$PERP_SEX <-as.factor(nypd.m$PERP_SEX)
nypd.m$VIC_SEX <-as.factor(nypd.m$VIC_SEX)
nypd.m$STATISTICAL_MURDER_FLAG <-as.factor(nypd.m$STATISTICAL_MURDER_FLAG)
nypd.m$PERP_AGE_GROUP <-as.factor(nypd.m$PERP_AGE_GROUP)
nypd.m$PERP_RACE <-as.factor(nypd.m$PERP_RACE)
nypd.m$VIC_AGE_GROUP <-as.factor(nypd.m$VIC_AGE_GROUP)
nypd.m$VIC_RACE <-as.factor(nypd.m$VIC_RACE)

```

The below model attempts to show if location description is significant predictor  
of a shooting being a murder

```{r fig.width=15, fig.height=9}
m1<-glm(STATISTICAL_MURDER_FLAG ~ LOCATION_DESC, data = nypd.m, family = "binomial")
summary(m1)
```

The above model shows that no particular location description is statistically 
significant when trying to predict if a shooting was a murder or not. P values
are too high.

The below model attempts to show if perp age group is significant predictor  
of a shooting being a murder

```{r fig.width=15, fig.height=9}
m2<-glm(STATISTICAL_MURDER_FLAG ~ PERP_AGE_GROUP, data = nypd.m, family = "binomial")
summary(m2)
```

According to the model, perp age group is a significant predictor of a shooting 
not being a murder. The model shows that they older the age group, the less likely
the shooting is to be a murder.


```{r fig.width=15, fig.height=9}
m2<-glm(STATISTICAL_MURDER_FLAG ~VIC_AGE_GROUP, data = nypd.m, family = "binomial")
summary(m2)
```

```{r fig.width=15, fig.height=9}
m3<-glm(STATISTICAL_MURDER_FLAG ~ BORO, data = nypd.m, family = "binomial")
summary(m3)
```

The above data model shows the Bronx, Brooklyn, and Manhattan are statistically
significant when predicting murders.


## Potential Bias in Analysis

When setting out on this analysis I initially assumed that younger folks were more
likely to be shooters than older folks and while this might be the case for ages
65+, this was not necessarily the case for ages 25-44. A look at the numbers 
mitigated these biases. That being said people ages 65+ were significantly less
likely to be shooters than those in the younger age groups.

As mentioned before another potential source of bias is resulting from
from the omission of a large amount of NA values when analyzing the perp data.
Perps who were not caught could potentially be over represented in a certain age
group which has the potential to skew the data one way or another. In this event,
we cannot be as conclusive with perp data as we are with victim data but the analysis
was still worthwhile. 


## Conclusion

Given our analysis above it seems that shooting location and age possibly tell
a story of proximity. Younger shooters and victims seem to both be primarily 
be in public housing and apartment buildings. This could imply that younger 
shooters tend to shoot younger victims more than they do older victims.


We also used 3 logistic regression modes to show the potential impact of location
description, perp/victim, and boro on the shooting being classified as a murder.

The models indicated that location description is not a good predictor of murder
and that perp/vic age group, and boro are statistically significant factors.

People in the <18-44 age group are more likely to be shot and murdered, and 
shootings that occur in the Bronx, Brooklyn, and Manhattan are more likely to be 
murders.










